\chapter{Theoretical Foundations: Heterogeneous Forest-Flow for ICU Trajectories}

This chapter establishes the mathematical and modelling foundations for the project; Chapters~\ref{ch:design} and~\ref{ch:implementation} translate those foundations into system design and implementation, respectively.

%==============================================================================
\section{Problem Formulation}
%==============================================================================

ICU trajectories derived from MIMIC-III \cite{johnson2016mimic} are modelled
through the preprocessing contract defined in
section~\ref{sec:mimic-preprocess-contract} (Design chapter), aligned with
MIMIC-Extract representation conventions \cite{wang2020mimicextract}.
For patient $i$, let:
\begin{itemize}\tightlist
    \item $\mathbf{s}_i \in \mathbb{R}^{d_s}$: static covariates (demographics/admission context),
    \item the dynamic state at hour $t$ be the tuple
    $\mathbf{x}_{i,t}=\bigl(\mathbf{x}_{i,t}^{\mathrm{cont}},\mathbf{x}_{i,t}^{\mathrm{bin}},\mathbf{x}_{i,t}^{\mathrm{cat}}\bigr)$:
    continuous features $\mathbf{x}_{i,t}^{\mathrm{cont}}\in\mathbb{R}^{d_{\mathrm{cont}}}$,
    binary features $\mathbf{x}_{i,t}^{\mathrm{bin}}\in\{0,1\}^{d_{\mathrm{bin}}}$,
    and categorical features with components $x_{i,t,j}^{\mathrm{cat}}\in\mathcal{C}_j$ for finite label sets $\mathcal{C}_j$, $j=1,\ldots,d_{\mathrm{cat}}$,
    \item $T_i$: observed trajectory length.
\end{itemize}
Let $d_x=d_{\mathrm{cont}}+d_{\mathrm{bin}}+d_{\mathrm{cat}}$ denote the total number of dynamic features (a fixed ordering is assumed for sequential generation in section~\ref{sec:hs3f}).

Clinically, $\mathbf{x}_{i,t}$ is mixed-type. During training, the same state is encoded numerically so that regressors and classifiers can be fit in a common pipeline. Continuous variables stay real-valued, binary variables are represented on $\{0,1\}$, and categorical variables are mapped to finite label sets. This categorical mapping is used partly for compute efficiency: compact integer class indices let tree classifiers operate directly on bounded label spaces, avoid one-hot expansion for high-cardinality columns, and reduce both memory footprint and per-fit training cost in repeated sweep runs. Where a single vector symbol appears below, it denotes the coordinates transported by continuous flow heads (or their encoded counterparts), not a claim that every variable is intrinsically continuous.

The joint law of the trajectory factorises as:
\begin{equation}
p(\mathbf{x}_{i,0:T_i} \mid \mathbf{s}_i)
= p(\mathbf{x}_{i,0} \mid \mathbf{s}_i)\prod_{t=1}^{T_i}p(\mathbf{x}_{i,t}\mid \mathbf{x}_{i,t-1}, \mathbf{s}_i).
\end{equation}
Each factor is shorthand for the appropriate product of conditional densities (continuous coordinates) and mass functions (discrete coordinates) consistent with the chosen factorisation (e.g.\ feature-wise in HS3F).

A two-stage factorisation with two coupled generators (hour-0 and transition) is adopted:
\begin{enumerate}\tightlist
    \item an IID hour-0 model for the joint initial vector $p(\mathbf{s},\mathbf{x}_{0})$,
    \item a conditional autoregressive model for transitions $p(\mathbf{x}_{t}\mid\mathbf{x}_{t-1},\mathbf{s})$.
\end{enumerate}
In this report, these views are used consistently as equivalent decompositions: with $\mathbf{s}$ observed from the cohort distribution, modelling $p(\mathbf{s},\mathbf{x}_0)$ is implemented as sampling $\mathbf{s}\sim p_{\text{data}}(\mathbf{s})$ and then modelling $p(\mathbf{x}_0\mid\mathbf{s})$.

The fixed feature ordering and this two-stage decomposition are design choices motivated by both modelling fidelity and implementation tractability. A fixed ordering gives a deterministic interface for sequential heads and reproducible train/sample behaviour across seeds. Separating hour-0 from transitions prevents the model from fabricating a pseudo-lag for $t=0$, keeps initial-state learning tied to real admission context, and avoids injecting unnecessary autoregressive noise at sequence start.
Thus the initial timestep need not be represented as if it were a lagged copy of a non-existent prior state.
The literature positioning of Forest-Flow and HS3F relative to other tabular and sequential generators (CTGAN, TimeGAN, TabDDPM, TabSyn, GReaT, REaLTabFormer) is discussed in the Background chapter and is not repeated here.
In plain terms, this formal setup separates initialization from transitions so the model can learn hour-0 structure and hour-to-hour dynamics with fewer avoidable artefacts.

%==============================================================================
\section{Flow Matching Foundation}
%==============================================================================

\subsection{Conditional Flow Matching}

Throughout this chapter, $\tau \in [0,1]$ denotes flow time (distinct from the physical hour index $t$ used in section~\ref{sec:hs3f} and the Problem Formulation above).

Flow Matching \cite{lipman2023flow} learns a time-indexed vector field that transforms noise to data on a real vector space aligned with continuously modelled coordinates (encoded as needed).
For such a target vector $\mathbf{x}$, Gaussian noise $\mathbf{z}\sim\mathcal{N}(0,I)$, and $\tau\sim\mathrm{Unif}[0,1]$:
\begin{equation}
\mathbf{x}_\tau = (1-\tau)\mathbf{z} + \tau\mathbf{x}.
\end{equation}

The supervised target field for independent conditional flow matching \cite{tong2024cfm} is:
\begin{equation}
\mathbf{u}_\tau = \mathbf{x} - \mathbf{z},
\end{equation}
and the model minimises:
\begin{equation}
\mathcal{L} = \mathbb{E}_{(\mathbf{x},\mathbf{c}),\mathbf{z},\tau}\left[\|\mathbf{v}_{\theta}(\mathbf{x}_\tau,\mathbf{c}) - \mathbf{u}_\tau\|_2^2\right],
\end{equation}
where $\mathbf{c}$ is the condition vector.
In plain terms, the model is trained to predict the direction from noise toward realistic data while holding context fixed, so generation becomes a controlled denoising trajectory rather than direct one-shot sampling.

\subsection{Clean-Condition Principle}

In conditional flow matching for autoregressive rollouts, noise is injected only into target coordinates routed to continuous flow heads. Categorical components are excluded from this interpolation in the implementation. Binary components, however, are currently transported as quantitative coordinates and then rounded at decode time to recover $\{0,1\}$ values.

This modelling convenience is important later: it creates a mechanism for out-of-support intermediate values and contributes to range and fidelity errors reported in Chapter~7 (see section~\ref{sec:sweep_quality}).
Condition features (static covariates and lagged dynamics) remain unperturbed during training:
\begin{equation}
\mathrm{input} = [\mathbf{x}_\tau^{\text{noisy}} \mid \mathbf{c}_{\text{clean}}].
\end{equation}
Here $\mathbf{x}_\tau^{\text{noisy}}$ denotes the noisy coordinates routed through flow regressors (continuous plus binary when typed quantitative); $\mathbf{c}_{\text{clean}}$ carries clean encoded context including lagged mixed-type information.
This choice is intended to support stable autoregressive conditioning and is consistent with the Forest-Flow family design \cite{jolicoeur2023forest}.
In plain terms, only the value being generated is noised while context stays clean, which improves rollout stability but leaves binary rounding as a known source of later fidelity errors.

%==============================================================================
\section{Tree-Based Approximation of the Vector Field}
%==============================================================================

Instead of neural networks, each time level is approximated with XGBoost regressors \cite{chen2016xgboost}.
Let $\{t_k\}_{k=0}^{n_t-1}$ with $t_k = k/n_t$.
For each $t_k$, the model fits regressors $\hat{\mathbf{v}}_{t_k}$ approximating the target field $\mathbf{u}_{t_k}$.

For Forest-Flow, sampling integrates forward in flow time from noise ($\tau=0$) to data ($\tau=1$), using the update:
\begin{equation}
\mathbf{x}_{k+1} = \mathbf{x}_{k} + h\hat{\mathbf{v}}_{t_k}(\mathbf{x}_{k}, \mathbf{c}), \quad k=0,\ldots,n_t-1,\; h=1/n_t,
\end{equation}
with $\mathbf{x}_0\sim\mathcal{N}(0,I)$ and $\mathbf{x}_{n_t}$ taken as the decoded sample.
In plain terms, this replaces one large difficult mapping with many small tree-based update steps that are CPU-friendly but can accumulate numerical error when step quality is weak.

%==============================================================================
\section{Heterogeneous Sequential Feature Modelling (HS3F)}
\label{sec:hs3f}
%==============================================================================

Flow fields are naturally continuous; categorical outputs are discrete.
HS3F \cite{akazan2024hs3f} addresses this mismatch by splitting generation by feature type and generating features sequentially.

\subsection{Feature-Wise Autoregressive Factorisation}

Given condition $\mathbf{c}_t=[\mathbf{s},\mathbf{x}_{t-1}]$, HS3F uses:
\begin{equation}
p(\mathbf{x}_t \mid \mathbf{c}_t)=\prod_{j=1}^{d_x}p(x_{t,j}\mid \mathbf{c}_t, x_{t,1:j-1}).
\end{equation}
Here $x_{t,1:j-1}$ denotes the already-generated hour-$t$ features preceding index $j$ in the fixed ordering (when $j=1$, this set is empty).
In plain terms, HS3F generates one feature at a time so each new value can depend on both patient context and features already generated at that same hour.

\subsection{Continuous vs. Categorical Heads}

For each feature $j$ in the fixed order (spanning continuous, binary, and multi-class categorical coordinates):
\begin{itemize}\tightlist
    \item If $j$ is routed to a quantitative head (continuous coordinates and, in the current implementation, binary coordinates): train flow-matching regressors over $t_k$.
    \item If $j$ is categorical (\code{'c'} in feature typing, typically multi-class): train an XGBoost classifier and sample from $\hat{p}(x_{t,j}=\cdot\mid \mathbf{c}_t, x_{t,1:j-1})$.
\end{itemize}

Gradient-boosted multiclass learners require categorical outcomes as contiguous integer class indices.
When training data are subsampled, the empirical support of categories may differ from the full corpus; consistent bookkeeping between the fitted model and the sampling map is needed in this pipeline to avoid index--label mismatches.
In plain terms, routing by type avoids forcing multi-class categories through continuous flow arithmetic, but it introduces extra label-mapping logic that must stay consistent at train and sample time.

\subsection{Numerical Integration Choice}

For continuous HS3F dimensions, the conditional vector field is integrated numerically along flow time. Forest-Flow uses first-order Euler updates \cite{jolicoeur2023forest}. HS3F defaults to fourth-order Runge--Kutta (RK4) for continuous heads \cite{akazan2024hs3f}.

RK4 has lower local truncation error per step under standard smoothness assumptions ($O(h^5)$ versus $O(h^2)$ for Euler). Euler is cheaper and matches the discrete grid used by Forest-Flow. The implementation mapping and solver settings are given in Chapter~6.
In plain terms, RK4 is chosen in HS3F to improve continuous-dimension integration accuracy, trading extra compute for potentially better stability in long rollouts.

%==============================================================================
\section{Computational Complexity}
%==============================================================================

Let $C_{\text{xgb}}$ denote average cost of one XGBoost fit call.

\paragraph{Forest-Flow (all continuous targets):}
\begin{equation}
\mathcal{O}(n_t \cdot d_x \cdot n_{\text{noise}} \cdot C_{\text{xgb}})
\end{equation}
where $n_{\text{noise}}$ (the number of noise replications per training row) controls the Monte-Carlo resolution of the flow target, for training, with sampling complexity roughly:
\begin{equation}
\mathcal{O}(n_{\text{synth}} \cdot n_t \cdot d_x \cdot C_{\text{pred}}).
\end{equation}

\paragraph{HS3F (mixed targets):}
If $d_q$ target dimensions are routed to quantitative heads (continuous plus binary under current typing) and $d_g$ are categorical classifier heads:
\begin{equation}
\mathcal{O}(n_t \cdot d_q \cdot n_{\text{noise}} \cdot C_{\text{xgb-reg}}
+ d_g \cdot C_{\text{xgb-clf}}).
\end{equation}

When categorical proportion is high, HS3F can avoid treating those coordinates as continuous flows and can be cheaper than an all-quantitative baseline when categorical classifier cost is small relative to $n_t$ flow-regressor fits \cite{akazan2024hs3f}.
In plain terms, the runtime cost is dominated by how many flow regressors are trained, which is why mixed-type routing can be both a fidelity choice and a compute-efficiency choice.
