\chapter{Design Specification and Current System Architecture}

%==============================================================================
\section{Data Pipeline and Representation}
%==============================================================================

\subsection{Source and Canonical Input}

The operational pipeline starts from hourly ICU tables derived from MIMIC-III \cite{johnson2016mimic}
through a MIMIC-Extract style workflow \cite{wang2020mimicextract}. The canonical training input for autoregressive modeling is \code{packages/data/processed/autoregressive\_data.csv}.
Hour-0 training uses \code{packages/data/processed/hour0\_data.csv}.

\subsection{Autoregressive Transformation}

Let $\mathcal{S}$ be static columns and $\mathcal{X}$ be dynamic columns.
With lag-1 engineering, each row at time $t$ uses:
\begin{align}
\text{target}_t &= \mathbf{x}_t \in \mathbb{R}^{|\mathcal{X}|}, \\
\text{condition}_t &= [\mathbf{s}, \mathbf{x}_{t-1}] \in \mathbb{R}^{|\mathcal{S}|+|\mathcal{X}|}.
\end{align}

\paragraph{Lag-0 and missing lag cells (implemented).}
The autoregressive table is built in \code{packages/data/autoregressive.py} by sorting on \code{(subject\_id, hours\_in)} and, for each dynamic column, computing \code{groupby(subject\_id)[col].shift(lag)} with \code{lag=1}.
\textbf{Lag-0 (no in-group history).} The first row of each \code{subject\_id} has no prior row in that group, so for each dynamic column \code{col} the lag feature \code{col\_lag1} is \code{NaN} before filling.
\textbf{Fill strategy and value.} When \code{fill\_strategy="special"}, the implementation assigns \code{df\_lagged.fillna(fill\_value)} to the entire lagged-column block.
The entry point \code{scripts/01\_preprocess\_data.py} calls \code{prepare\_autoregressive\_data} with \code{fill\_strategy="special"} and \code{fill\_value=-1.0} (matching the function defaults for those parameters), so every lag-column \code{NaN} becomes the float sentinel \code{-1.0}.
The same fill applies to \code{NaN}s arising when $t-1$ exists but the source value at $t-1$ was missing: \code{shift} propagates \code{NaN}, which is then indistinguishable from lag-0 after filling.

\paragraph{Rationale and risks.}
The sentinel keeps a dense, rectangular lag feature matrix so hour~0 rows remain in the autoregressive CSV and need not be dropped for lack of history; the IID hour-0 pipeline can still model the initial state separately.
Risks: \code{-1.0} is not a reserved missing code in the raw feature space---it can collide with legitimate negative measurements or encodings that use $-1$---and one code cannot separate structural ``no prior timestep'' from ``prior observation missing'' in the flat table.

\paragraph{Mitigation (recommended checks).}
Treat the choice as an empirical assumption.
As a sensitivity analysis or ablation, rerun with other supported options in code (\code{fill\_strategy} \code{zero} or \code{mean}, or \code{special} with a \code{fill\_value} chosen per feature scale) and compare validation or short rollouts, with particular attention to variables where $-1$ is plausible.
Outcomes remain dataset- and model-dependent, and improvement is not guaranteed.

\subsection{Column Roles Persisted as Metadata}

The preprocessor artifact stores:
\begin{itemize}\tightlist
    \item \code{target\_cols} (semantic generation outputs),
    \item \code{condition\_cols} (semantic conditioning inputs),
    \item \code{target\_indices} and \code{condition\_indices} in transformed space.
\end{itemize}

This design persists both semantic columns and transformed indices because transformed column order is type-grouped (numeric, binary, categorical), not semantic target-then-condition order.

%==============================================================================
\section{Tabular Preprocessor Design}
%==============================================================================

\subsection{Feature-Type Identification}

Feature typing is centralized in \code{packages/data/feature\_utils.py} and consumed by \code{scripts/01b\_fit\_preprocessor.py}.
Columns are grouped into:
\begin{itemize}\tightlist
    \item numeric continuous,
    \item binary 0/1,
    \item categorical,
    \item integer-like numeric (for inverse rounding policy).
\end{itemize}

\subsection{Transform Semantics (Current)}

The current \code{TabularPreprocessor} is intentionally simple:
\begin{enumerate}\tightlist
    \item Numeric: pass-through (no normalization).
    \item Binary: pass-through in transform; round/clip policies in inverse path.
    \item Categorical: label encoding to integer IDs.
\end{enumerate}

This design differs from older one-hot/min-max variants and aligns with the tree-based backend, where strict scaling is not required for split-based learners \cite{chen2016xgboost}.

\subsection{Streaming Fit for Full-Dataset Coverage}

\code{scripts/01b\_fit\_preprocessor.py} performs chunked fitting on the full autoregressive CSV via repeated \code{partial\_fit()} calls followed by \code{finalize\_fit()}, then persists the artifact at \code{models/preprocessor\_full.pkl}.

The design goal is to capture rare categorical values and robust numeric ranges before any subsampled training.

\subsection{Inverse Transform Contract}

\code{inverse\_transform()} reconstructs a human-readable DataFrame by:
\begin{itemize}\tightlist
    \item decoding categorical IDs back to labels,
    \item applying binary/integer rounding rules,
    \item preserving numeric values in raw scale.
\end{itemize}

This decode step is part of the intended export path for downstream CSV outputs and clinical interpretability, even though model training happens in encoded numeric space.

\subsection{Index-Based Split as a Design Constraint}

Because transform output order is type-grouped:
\begin{equation}
\text{transformed\_order} = [\text{numeric},\text{binary},\text{categorical}],
\end{equation}
target and condition blocks may become non-contiguous in transformed coordinates.

Therefore training scripts use:
\begin{equation}
X_{\text{target}} = X[:, \text{target\_indices}], \quad
X_{\text{condition}} = X[:, \text{condition\_indices}],
\end{equation}
instead of naive prefix slicing.

%==============================================================================
\section{Generator Backbone Architecture}
%==============================================================================

\subsection{System Pipeline Overview}

Figure~\ref{fig:pipeline-overview} shows the end-to-end pipeline from raw MIMIC-III ICU tables to evaluated synthetic output. The two-stage architecture is explicit: the hour-0 IID model and the autoregressive transition model are trained independently on their respective data slices, and synthetic trajectories are assembled by chaining their outputs.

% ---------------------------------------------------------------------------
% Figure 1: Two-stage pipeline overview
% ---------------------------------------------------------------------------
\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=0.88, transform shape,
  node distance = 6mm and 12mm,
  box/.style = {rectangle, draw, rounded corners=2pt,
                minimum width=32mm, minimum height=8mm,
                text centered, font=\small},
  data/.style = {box, fill=blue!10},
  proc/.style = {box, fill=orange!15},
  model/.style= {box, fill=green!15},
  eval/.style = {box, fill=red!10},
  arr/.style  = {-Stealth, thick},
  grp/.style  = {draw=gray!60, dashed, rounded corners=4pt,
                 inner sep=4pt, fill=gray!5}
]

% --- Data preparation column (left) ---
\node[data]  (mimic)   {MIMIC-III ICU tables};
\node[proc,  below=of mimic]  (p00)    {\texttt{00\_prepare}};
\node[proc,  below=of p00]    (p01)    {\texttt{01\_preprocess}};
\node[proc,  below=of p01]    (p01b)   {\texttt{01b\_fit\_preprocessor}};

% --- Hour-0 branch (middle column) ---
% Anchored at the height of p00 so both training nodes are at the same level
\node[model, right=24mm of p00]  (h0train)  {Train hour-0 IID model};
\node[model, below=of h0train]   (h0gen)    {Sample initial states $\hat{\mathbf{x}}_0$};

% --- Autoregressive branch (right column, same top height as h0 branch) ---
\node[model, right=20mm of h0train] (artrain)  {Train AR model};
\node[model, below=of artrain]      (argen)    {Autoregressively roll-out $\hat{\mathbf{x}}_{1:T}$};

% --- Merge: centred between h0gen and argen ---
\node[proc] at ($(h0gen.south)!0.5!(argen.south) + (0,-16mm)$) (merge)
      {Assemble synthetic cohort CSV};

% --- Evaluation stack (below merge) ---
\node[eval, below=of merge] (tstr)   {Patient-level TSTR (mortality, LOS)};
\node[eval, below=of tstr]  (qual)   {KS / Correlation / Range violations};
\node[eval, below=of qual]  (priv)   {DCR + DOMIAS-style MIA};

% --- Arrows: data prep ---
\draw[arr] (mimic) -- (p00);
\draw[arr] (p00)   -- (p01);
\draw[arr] (p01)   -- (p01b);

% --- Arrows: data prep → branches ---
\draw[arr] (p00)   -- (h0train);
\draw[arr] (p01b)  -- (artrain);

% --- Arrows: within branches ---
\draw[arr] (h0train) -- (h0gen);
\draw[arr] (artrain) -- (argen);

% --- Arrows: converge to merge ---
\draw[arr] (h0gen.south) -- (merge.north west);
\draw[arr] (argen.south) -- (merge.north east);

% --- Arrows: evaluation ---
\draw[arr] (merge) -- (tstr);
\draw[arr] (tstr)  -- (qual);
\draw[arr] (qual)  -- (priv);

% --- Grouping boxes ---
\begin{scope}[on background layer]
  \node[grp, fit=(p00)(p01)(p01b), label=above:{\scriptsize Data preparation}] {};
  \node[grp, fit=(h0train)(h0gen), label=above:{\scriptsize Hour-0 model}] {};
  \node[grp, fit=(artrain)(argen), label=above:{\scriptsize Autoregressive model}] {};
  \node[grp, fit=(tstr)(qual)(priv), label=below:{\scriptsize Evaluation stack}] {};
\end{scope}

\end{tikzpicture}
\caption{End-to-end pipeline. MIMIC-III ICU tables feed two parallel training paths: an IID hour-0 model (generating initial patient states) and an autoregressive transition model (generating subsequent hourly observations). Their outputs are chained into a synthetic cohort CSV, which is passed through the three-axis evaluation stack.}
\label{fig:pipeline-overview}
\end{figure}

\subsection{ForestFlow Architecture}

Figure~\ref{fig:forestflow-arch} shows the ForestFlow backbone. At training time, for each of $n_t$ discrete time levels, a noise-augmented version of the target is constructed by interpolating between Gaussian noise $\mathbf{z}$ and the real target $\mathbf{x}$, then concatenated with the clean condition vector. An XGBoost regressor is fitted to predict the flow direction $\mathbf{u}_t = \mathbf{x} - \mathbf{z}$. At generation time, Euler integration steps backward from noise to data using the learned regressors.

% ---------------------------------------------------------------------------
% Figure 2: ForestFlow backbone
% ---------------------------------------------------------------------------
\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=0.88, transform shape,
  node distance = 5mm and 12mm,
  box/.style  = {rectangle, draw, rounded corners=2pt,
                 minimum width=30mm, minimum height=8mm,
                 text centered, font=\small},
  blue/.style = {box, fill=blue!10},
  orange/.style={box, fill=orange!15},
  green/.style ={box, fill=green!15},
  red/.style   ={box, fill=red!10},
  arr/.style   = {-Stealth, thick},
  darr/.style  = {-Stealth, thick, dashed},
  grp/.style   = {draw=gray!60, dashed, rounded corners=4pt,
                  inner sep=4pt, fill=gray!5}
]

% ---- Training path (left column) ----
\node[blue]   (noise)  {Sample $\mathbf{z}\sim\mathcal{N}(0,I)$};
\node[blue,   below=of noise]   (interp) {Interpolate: $\mathbf{x}_t=(1{-}t)\mathbf{z}+t\mathbf{x}$};
\node[orange, below=of interp]  (concat) {Concat $[\mathbf{x}_t^{\text{noisy}} \mid \mathbf{c}_{\text{clean}}]$};
\node[green,  below=of concat]  (xgb)    {XGBoost regressor fits $\hat{\mathbf{v}}_{t_k}$};
\node[orange, below=of xgb]     (loss)   {Loss: $\|\hat{\mathbf{v}}_{t_k}-(\mathbf{x}-\mathbf{z})\|_2^2$};

% ---- Time loop annotation ----
\node[right=5mm of xgb, font=\scriptsize, text=gray] (timeann)
      {$\times\; n_t$ time levels};

% ---- Sampling path (right column) ----
\node[blue,   right=28mm of noise]   (znoise) {Start: $\hat{\mathbf{x}}_1\sim\mathcal{N}(0,I)$};
\node[green,  below=of znoise]       (pred)   {Predict $\hat{\mathbf{v}}_{t_k}(\hat{\mathbf{x}}_k,\mathbf{c})$};
\node[orange, below=of pred]         (euler)  {Euler step: $\hat{\mathbf{x}}_{k-1}=\hat{\mathbf{x}}_k+h\hat{\mathbf{v}}_{t_k}$};
\node[blue,   below=of euler]        (out)    {Output: $\hat{\mathbf{x}}_0$ (decoded)};

% ---- Arrows: training ----
\draw[arr] (noise)  -- (interp);
\draw[arr] (interp) -- (concat);
\draw[arr] (concat) -- (xgb);
\draw[arr] (xgb)    -- (loss);

% ---- Arrows: sampling ----
\draw[arr] (znoise) -- (pred);
\draw[arr] (pred)   -- (euler);
\draw[arr] (euler)  to[out=0,in=0,looseness=2]
           node[right,font=\scriptsize]{$k{=}n_t{\to}1$} (pred);
\draw[arr] (euler)  -- (out);

% ---- Training → Sampling arrow ----
\draw[darr] (xgb.east) -- node[above,font=\scriptsize]{stored regressors} (pred.west);

% ---- Group boxes ----
\begin{scope}[on background layer]
  \node[grp, fit=(noise)(interp)(concat)(xgb)(loss),
        label=above:{\scriptsize Training}] {};
  \node[grp, fit=(znoise)(pred)(euler)(out),
        label=above:{\scriptsize Sampling (Euler)}] {};
\end{scope}

\end{tikzpicture}
\caption{ForestFlow backbone. \textit{Training}: for each time level $t_k$, the target is interpolated with Gaussian noise and a regressor is fitted to predict the flow direction. \textit{Sampling}: starting from noise, $n_t$ first-order Euler steps integrate the learned vector field back to data space. All target dimensions are treated as continuous flows.}
\label{fig:forestflow-arch}
\end{figure}

\subsection{HS3F Architecture}

Figure~\ref{fig:hs3f-arch} shows the HS3F backbone. The key difference from ForestFlow is sequential feature-wise generation: features are produced one at a time in a fixed order, each conditioned on previously generated features. Continuous features use the same flow-matching regressor path as ForestFlow; categorical (multi-class) features are generated by an XGBoost classifier trained directly on the class label.

% ---------------------------------------------------------------------------
% Figure 3: HS3F backbone
% ---------------------------------------------------------------------------
\begin{figure}[htbp]
\centering
\begin{tikzpicture}[scale=0.78, transform shape,
  node distance = 5mm and 10mm,
  box/.style   = {rectangle, draw, rounded corners=2pt,
                  minimum width=30mm, minimum height=8mm,
                  align=center, font=\small},
  blue/.style  = {box, fill=blue!10},
  orange/.style= {box, fill=orange!15},
  green/.style = {box, fill=green!15},
  purple/.style= {box, fill=purple!10},
  red/.style   = {box, fill=red!10},
  arr/.style   = {-Stealth, thick},
  darr/.style  = {-Stealth, thick, dashed},
  grp/.style   = {draw=gray!60, dashed, rounded corners=4pt,
                  inner sep=4pt, fill=gray!5}
]

% ---- Shared input ----
\node[blue]   (cond)  {Condition $\mathbf{c} = [\mathbf{s},\, \mathbf{x}_{t-1}]$};

% ---- Feature loop head (router) ----
\node[orange, below=10mm of cond]  (router) {Feature $j$ router: type check};
\node[right=3mm of router, font=\scriptsize, text=gray]
     (loopann) {for $j=1,\ldots,d_x$};

% ---- Continuous branch ----
\node[green, below left=12mm and 12mm of router]  (flow)
     {Flow-matching regressor\\$\hat{v}_{t_k}(x_{t,j}\mid\mathbf{c},x_{t,<j})$};
\node[green, below=of flow] (rk4)
     {RK4 integration $\to \hat{x}_{t,j}$};

% ---- Categorical branch ----
\node[purple, below right=12mm and 12mm of router] (clf)
     {XGBoost classifier\\$\hat{p}(x_{t,j}\mid\mathbf{c},x_{t,<j})$};
\node[purple, below=of clf] (sample)
     {Sample $\hat{x}_{t,j}\sim\hat{p}(\cdot)$};

% ---- Accumulate ----
\node[blue, below=20mm of router]  (accum)
     {Append $\hat{x}_{t,j}$ to generated prefix};
\node[orange, below=of accum] (done)
     {All $d_x$ features generated: output $\hat{\mathbf{x}}_t$};

% ---- Arrows ----
\draw[arr] (cond)   -- (router);
\draw[arr] (router) -- node[left, font=\scriptsize]{continuous} (flow);
\draw[arr] (router) -- node[right,font=\scriptsize]{categorical} (clf);
\draw[arr] (flow)   -- (rk4);
\draw[arr] (clf)    -- (sample);
\draw[arr] (rk4)    -- (accum);
\draw[arr] (sample) -- (accum);
\draw[arr] (accum)  -- (done);
% Feedback: generated feature feeds back into condition for next j
\draw[darr] (accum.west) to[out=180,in=180,looseness=1.4]
            node[left,font=\scriptsize]{feeds $x_{t,<j+1}$} (router.west);

% ---- Group boxes ----
\begin{scope}[on background layer]
  \node[grp, fit=(flow)(rk4), label=left:{\scriptsize Continuous head}] {};
  \node[grp, fit=(clf)(sample),  label=right:{\scriptsize Categorical head}] {};
\end{scope}

\end{tikzpicture}
\caption{HS3F backbone. Features are generated sequentially in a fixed order. Continuous features are produced by a flow-matching regressor integrated with RK4; categorical features are produced by an XGBoost classifier with multinomial sampling. Each generated feature is appended to the conditioning prefix before the next feature is generated, capturing inter-feature dependencies.}
\label{fig:hs3f-arch}
\end{figure}

\subsection{Backbone Selection}

All major training entry points accept the flag \code{--model-type} with values \code{forest-flow} or \code{hs3f}; \code{hs3f} is the default.

\subsection{ForestFlow Backbone}

The \code{ForestFlow} class in \code{packages/models/forest\_flow.py} implements conditional flow matching with XGBoost regressors.
Core interface:
\begin{itemize}\tightlist
    \item \code{fit(X\_target, X\_condition, feature\_types)},
    \item \code{sample(n\_samples, X\_condition, random\_state)}.
\end{itemize}

Design details:
\begin{enumerate}\tightlist
    \item Time discretization into \code{nt} levels.
    \item Noise duplication control via \code{n\_noise}.
    \item Optional iterator path for memory-aware XGBoost training.
    \item Conditional input formed as \code{[noisy\_target | clean\_condition]}.
\end{enumerate}

Categorical typing logic is constrained by flow arithmetic: target dimensions that receive noise are forced to quantitative mode during training input assembly.

\subsection{HS3F Backbone}

The \code{HS3F} class in \code{packages/models/hs3f.py} follows heterogeneous sequential generation \cite{akazan2024hs3f}:
\begin{itemize}\tightlist
    \item continuous targets: flow-matching regressors,
    \item categorical targets: XGBoost classifiers with multinomial sampling,
    \item left-to-right feature generation conditioned on previously generated features.
\end{itemize}

Continuous heads integrate the learned flow in time; categorical heads apply label remapping so that, when training on subsets of the data, class indices used during fitting align with those emitted at sampling time.
This design is intended to mitigate the sparse-class-index failure mode common under subsampling.

\subsection{Hour-0 + Autoregressive Coupling}

The architecture is intentionally two-stage:
\begin{enumerate}\tightlist
    \item Hour-0 IID model learns initial states.
    \item Autoregressive model learns transitions conditioned on lagged dynamics and static covariates.
\end{enumerate}

This decomposition separates structural initialization from transition dynamics more explicitly than a single model that must infer both jointly.

%==============================================================================
\section{Sweep-Centric System Design}
%==============================================================================

\subsection{Unified Hyperparameter Protocol}

The sweep orchestrates paired training of the hour-0 and autoregressive models for each $(n_t, n_{\text{noise}})$ configuration so that synthetic rollouts use mutually consistent stage-one and stage-two parameters for that configuration.

\subsection{Always-On Transformed Cache}

Transformed autoregressive matrices and a schema-consistent real-quality reference sample are persisted under \code{models/sweep\_cache/transformed\_autoregressive/} to avoid repeating expensive transform work across experiment launches.

Entries are treated as valid only when a content signature matches the current source table and target/condition index specification; signature fields and reload rules are defined in the sweep implementation (\code{scripts/run\_sweep.py}) to prevent stale reuse after schema or split changes.

\subsection{Result Schema Stability}

Sweep outputs are written to a single append-only table with a fixed column contract so that partial reruns merge cleanly and historical rows remain comparable across code revisions.
The concrete column set is defined alongside the sweep driver implementation.

%==============================================================================
\section{Design Rationale and Trade-offs}
%==============================================================================

\subsection{Why Raw Numeric Pass-Through?}

Tree boosting is split-based and generally robust to monotonic scaling transformations \cite{chen2016xgboost}.
Removing mandatory min-max scaling is intended to simplify preprocessing and reduce opportunities for inverse-transform coupling errors.

\subsection{Why Keep Inverse Transform?}

Even with simplified encoding, inverse transformation remains important for:
\begin{itemize}\tightlist
    \item readable synthetic table export,
    \item downstream evaluator compatibility,
    \item sanity checks in clinical units.
\end{itemize}

\subsection{Why Multi-Backbone Support?}

Forest-Flow \cite{jolicoeur2023forest} remains a well-established baseline and compatibility path.
HS3F \cite{akazan2024hs3f} is aligned with heterogeneous mixed-type generation in the cited work and is the default backbone in the current implementation.

\subsection{Why Preserve Explicit Metadata Artifacts?}

Persisting split metadata and model type in pickled artifacts enables:
\begin{itemize}\tightlist
    \item backward-compatible loading in generation scripts,
    \item reproducible sweep reruns,
    \item traceable migration between architecture revisions.
\end{itemize}
