%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% 6CCS3PRJ Final Year Individual Project Report
% luke.day@kcl.ac.uk
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\documentclass[11pt]{informatics-report}
\usepackage{color}
\usepackage[square,comma,numbers]{natbib}
% Required for [noitemsep] to work
\usepackage{enumitem}

% Other Forest-Flow packages
\usepackage{amsmath,amssymb,amsthm}
% Custom commands
\newcommand{\R}{\mathbb{R}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Prob}{\mathbb{P}}
\newcommand{\given}{\mid}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{booktabs}
\usepackage{graphicx}
\graphicspath{{figures/}{../results/sweep_plots/}}
\usepackage{xcolor}
\usepackage{microtype}
\usepackage{tabularx}
\usepackage{array}
\usepackage{listings}%References
\lstdefinestyle{codeinline}{%
  basicstyle=\ttfamily,
  breaklines=true,
  columns=fullflexible,
}
\newcommand{\code}[1]{\lstinline[style=codeinline]!#1!}
% TikZ for architecture diagrams
\usepackage{tikz}
\usetikzlibrary{shapes.geometric,arrows.meta,positioning,fit,backgrounds,calc}
% Pandoc-compatible tight list (removes extra spacing in itemize/enumerate)
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Front Matter - project title, name, supervisor name and date
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Synthetic ICU Trajectory Generation\\\vspace{0.2cm}via Heterogeneous Tree-Based Flow Matching}
\author{Lian Matsuo\\Programme of Study: Artificial Intelligence}
\studentID{K23060243}
\supervisor{David Watson}

\date{\today}

\abstractFile{FrontMatter/abstract.tex}
\ackFile{FrontMatter/acknowledgements.tex} %Remove line if you do not want acknowledgements

\begin{document}
\setlength{\emergencystretch}{3em}
\createFrontMatter
\onehalfspacing
\tableofcontents
\doublespacing

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Report Content
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% You can write each chapter directly here or in a separate .tex file and use the include command.

\chapter{Introduction}

Healthcare data is difficult to share and model reliably at scale. Electronic Health Records (EHRs) combine mixed numerical and categorical variables, sparse and irregular dependencies, long-tailed outcomes, and domain rules that must be respected. Data-governance constraints under frameworks such as UK GDPR and the Data Protection Act 2018 further restrict access \cite{dataProtectionAct2018,icoResearchSafeguards}, while rare diseases and minority subpopulations create severe data scarcity and class imbalance. These weaknesses in real-world data availability and coverage are a core reason synthetic data generation is actively studied in healthcare.

Synthetic tabular generation can mitigate privacy and scarcity constraints, but common families still face limitations on mixed-type clinical data. GANs can underrepresent rare patterns, VAEs can oversmooth heterogeneous structure, and neural diffusion models often demand higher compute while requiring careful handling of categorical constraints.

Recent tree-based flow-matching work, especially ForestFlow \cite{jolicoeur2023forest} and HS3F \cite{akazan2024hs3f}, shows that gradient-boosted trees can approximate flow fields effectively for mixed-type tabular synthesis. This is directly relevant to ICU trajectories because tree partitions better match tabular heterogeneity and enable CPU-feasible training.

This project applies tree-based flow matching to sequential ICU patient trajectories derived from the MIMIC-III clinical database \cite{johnson2016mimic}. Unlike prior work that evaluates ForestFlow and HS3F on static, single-row tabular benchmarks, this project extends the approach to an autoregressive setting: a two-stage architecture where an independent and identically distributed (IID) hour-0 model generates initial patient states and a conditional model generates subsequent hourly transitions given lagged dynamics and static covariates. Two model backbones are implemented and compared: ForestFlow (all-continuous flow) and HS3F (heterogeneous sequential generation with separate continuous and categorical heads).

The objective is to investigate whether tree-based flow matching can produce synthetic ICU trajectories with useful fidelity, downstream utility, and privacy diagnostics while training efficiently on CPUs. The focus is technical feasibility and limitation-mapping rather than deployment claims.

Evaluation follows a multi-axis protocol: (1) \emph{fidelity} via Kolmogorov--Smirnov statistics, correlation preservation, and clinical range violation rates; (2) \emph{utility} via multi-task Train-on-Synthetic, Test-on-Real (TSTR) evaluation covering patient-level mortality prediction and ICU length-of-stay (LOS) prediction; and (3) \emph{privacy} via Distance-to-Closest-Record diagnostics and overfitting-based membership inference attacks (DOMIAS-style).

%==============================================================================
\section{Research Claim}
%==============================================================================

\noindent This dissertation argues that tree-based flow matching, originally formulated for static tabular data, can be adapted to sequential ICU trajectory synthesis on MIMIC-III through a two-stage autoregressive factorisation. Results from the sweep and matched backbone comparisons indicate technical feasibility: the generated data retains a measurable predictive signal and avoids exact-record copying in DCR checks. However, those same results also show practical limitations, particularly weak minority-class utility and high clinical range violations, so operational release readiness is not established.

%==============================================================================
\section{Aim and Objectives}
%==============================================================================

\paragraph{Aim:}
The aim of this project is to develop and evaluate a tree-based flow-matching framework for synthesising sequential ICU patient trajectories from the MIMIC-III clinical database \cite{johnson2016mimic}. The project assesses whether this approach can preserve useful structure across fidelity, downstream utility, privacy diagnostics, and computational efficiency in a reproducible research setting.

\paragraph{Objectives:}
This aim is operationalised through four concrete objectives:

\begin{description}[leftmargin=0cm,labelsep=0.6em]
    \item[Domain adaptation:] Extend ForestFlow \cite{jolicoeur2023forest} and HS3F \cite{akazan2024hs3f}, tree-based flow-matching models originally evaluated on static tabular benchmarks, to sequential clinical data via an autoregressive two-stage architecture comprising an IID hour-0 model for initial patient states and a conditional transition model for subsequent hourly dynamics.
    \item[Backbone implementation and comparison:] Implement both ForestFlow and HS3F within a unified training and evaluation interface so they can be compared under \textbf{identical} MIMIC preparation, data budgets, and evaluation code; the systematic hyperparameter grid (see below) is used to \textbf{map HS3F} over $(n_t, n_{\text{noise}})$, while the head-to-head evidence plan for all-continuous versus heterogeneous routing is a \textbf{pre-specified matched cell} (with CTGAN as a non-autoregressive reference), not a full per-cell multi-backbone matrix.
    \item[Multi-axis evaluation framework:] Develop a reproducible evaluation stack covering (i) distributional fidelity via Kolmogorov--Smirnov statistics, correlation preservation, and clinical range violation rates; (ii) downstream utility via multi-task Train-on-Synthetic, Test-on-Real (TSTR) evaluation covering patient-level mortality prediction and ICU length-of-stay (LOS) prediction; and (iii) privacy risk via Distance-to-Closest-Record diagnostics and DOMIAS-style membership inference attacks \cite{breugel2023domias}.
    \item[Systematic hyperparameter investigation:] Evaluate the effect of key flow-matching hyperparameters ($n_t$, $n_{\text{noise}}$) on synthetic data quality, utility, and computational cost via a reproducible, resumable grid sweep.
\end{description}

\paragraph{Research questions:}
The objectives above are evaluated through the following specific research questions:

\begin{description}[leftmargin=0cm,labelsep=0.6em]
    \item[Feasibility:] Can tree-based flow matching, originally evaluated on static tabular benchmarks, be extended to generate realistic sequential ICU trajectories via autoregressive conditioning?
    \item[Heterogeneous features (matched backbones):] At a fixed, shared hyperparameter cell with identical MIMIC cohort, training budget, and trajectory seeds, how do fidelity- and utility-relevant metrics compare between HS3F (explicit continuous/categorical heads) and ForestFlow (all-continuous flow) on the same mixed-type autoregressive tables? (The $8 \times 8$ sweep in this thesis \textbf{characterises HS3F} over $(n_t, n_{\text{noise}})$; it is not, and is not presented as, a grid in which every cell reruns all backbones. Evidence for the backbone question is the matched design in section~\ref{sec:backbone-comparison}.)
    \item[Utility and privacy:] Do synthetic trajectories preserve sufficient downstream predictive signal (TSTR) while maintaining acceptable privacy-risk indicators under proximity-based and attack-based diagnostics?
\end{description}

Throughout the report, a central distinction is maintained between \emph{research feasibility} and \emph{practical adequacy}: a configuration may generate coherent trajectories and transfer a predictive signal while still falling short of release or operational standards.

Although this project does not propose new generative algorithms, it contributes an \textbf{application-level integration} of existing tree-based flow-matching methods into a new domain (sequential clinical data) and a new problem structure (autoregressive trajectory generation).
The contribution lies in the system design, the multi-axis evaluation framework, and the empirical investigation of how these methods behave when adapted from static benchmarks to the challenges of temporal, mixed-type ICU data.

\include{Chapters/Background}
\include{Chapters/Requirements}
\include{Chapters/Body}
\include{Chapters/DesignSpecification}
\include{Chapters/Implementation}
\include{Chapters/Evaluation}
\include{Chapters/ProfessionalIssues}
\include{Chapters/Conclusion}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% References
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{plain}
\begin{thebibliography}{99}

\bibitem{johnson2016mimic}
Johnson, A. E. W., Pollard, T. J., Shen, L., Lehman, L.-w. H., Feng, M., Ghassemi, M.,
Moody, B., Szolovits, P., Celi, L. A., and Mark, R. G. (2016).
\textit{MIMIC-III, a freely accessible critical care database}.
Scientific Data, 3, 160035.

\bibitem{wang2020mimicextract}
Wang, S., McDermott, M. B. A., Chauhan, G., Ghassemi, M., Hughes, M. C., and Naumann, T. (2020).
\textit{MIMIC-Extract: A Data Extraction, Preprocessing, and Representation Pipeline for MIMIC-III}.
Proceedings of the ACM Conference on Health, Inference, and Learning (CHIL).

\bibitem{silva2012challenge}
Silva, I., Moody, B., Scott, D. J., Celi, L. A., and Mark, R. G. (2012).
\textit{Predicting In-Hospital Mortality of ICU Patients: The PhysioNet/Computing in Cardiology Challenge 2012}.
Computing in Cardiology, 39, 245--248.

\bibitem{lipman2023flow}
Lipman, Y., Chen, R. T. Q., Ben-Hamu, H., Nickel, M., and Le, M. (2023).
\textit{Flow Matching for Generative Modeling}.
International Conference on Learning Representations (ICLR).

\bibitem{jolicoeur2023forest}
Jolicoeur-Martineau, A., Fatras, K., and Kachman, T. (2024).
\textit{Generating and Imputing Tabular Data via Diffusion and Flow-based Gradient-Boosted Trees}.
Proceedings of the 27th International Conference on Artificial Intelligence and Statistics (AISTATS).

\bibitem{akazan2024hs3f}
Akazan, A.-C., Jolicoeur-Martineau, A., and Mitliagkas, I. (2024).
\textit{Generating Tabular Data Using Heterogeneous Sequential Feature Forest Flow Matching}.
arXiv:2410.15516.

\bibitem{chen2016xgboost}
Chen, T., and Guestrin, C. (2016).
\textit{XGBoost: A Scalable Tree Boosting System}.
Proceedings of the 22nd ACM SIGKDD International Conference on Knowledge Discovery and Data Mining.

\bibitem{esteban2018rcgan}
Esteban, C., Hyland, S. L., and R{\"a}tsch, G. (2018).
\textit{Real-valued (Medical) Time Series Generation with Recurrent Conditional GANs}.
International Conference on Learning Representations (ICLR).

\bibitem{xu2019ctgan}
Xu, L., Skoularidou, M., Cuesta-Infante, A., and Veeramachaneni, K. (2019).
\textit{Modeling Tabular Data Using Conditional GAN}.
Advances in Neural Information Processing Systems (NeurIPS).

\bibitem{yoon2019timegan}
Yoon, J., Jarrett, D., and van der Schaar, M. (2019).
\textit{Time-series Generative Adversarial Networks}.
Advances in Neural Information Processing Systems (NeurIPS).

\bibitem{sdvproject}
Patki, N., Wedge, R., and Veeramachaneni, K. (2016).
\textit{The Synthetic Data Vault}.
IEEE International Conference on Data Science and Advanced Analytics Workshops.
\newline
Project ecosystem: \url{https://github.com/sdv-dev}.
(Accessed: 20 April 2026).

\bibitem{agileBusinessConsortiumMoSCoW}
Agile Business Consortium.
\textit{DSDM Project Framework Handbook: Chapter 10 -- MoSCoW Prioritisation}.
\url{https://www.agilebusiness.org/dsdm-project-framework/moscow-prioritisation.html}.
(Accessed: 20 April 2026).

\bibitem{runesonHost2009CaseStudyGuidelines}
Runeson, P., and H{\"o}st, M. (2009).
\textit{Guidelines for conducting and reporting case study research in software engineering}.
Empirical Software Engineering, 14(2), 131--164.

\bibitem{shokri2017membership}
Shokri, R., Stronati, M., Song, C., and Shmatikov, V. (2017).
\textit{Membership Inference Attacks Against Machine Learning Models}.
IEEE Symposium on Security and Privacy, 3--18.

\bibitem{breugel2023domias}
van Breugel, B., Sun, H., Qian, Z., and van der Schaar, M. (2023).
\textit{Membership Inference Attacks against Synthetic Data through Overfitting Detection}.
Proceedings of The 26th International Conference on Artificial Intelligence and Statistics (AISTATS), PMLR 206, 3493--3514.

\bibitem{lautrup2024syntheval}
Lautrup, A. D., Hyrup, T., Zimek, A., and Schneider-Kamp, P. (2024).
\textit{SynthEval: A Framework for Detailed Utility and Privacy Evaluation of Tabular Synthetic Data}.
arXiv:2404.15821.

\bibitem{yao2025dcrdelusion}
Yao, Z., Kr{\v{c}}o, N., Ganev, G., and de Montjoye, Y.-A. (2025).
\textit{The DCR Delusion: Measuring the Privacy Risk of Synthetic Data}.
arXiv:2505.01524.

\bibitem{sdmetricsPrivacyDocs}
SDV project documentation (2026).
\textit{SDMetrics Privacy Metrics: DCRBaselineProtection and DCROverfittingProtection}.
\url{https://docs.sdv.dev/sdmetrics/data-metrics/privacy}.
(Accessed: 20 April 2026).

\bibitem{mimicGettingStarted}
MIMIC Project (2021).
\textit{Getting Started: Access}.
\url{https://mimic.physionet.org/gettingstarted/access}.
(Accessed: 20 April 2026).

\bibitem{physionetCredentialedLicense150}
MIT Laboratory for Computational Physiology.
\textit{PhysioNet Credentialed Health Data License 1.5.0}.
\url{https://physionet.org/about/licenses/physionet-credentialed-health-data-license-150}.
(Accessed: 20 April 2026).

\bibitem{dataProtectionAct2018}
UK Parliament (2018).
\textit{Data Protection Act 2018 (c. 12)}.
\url{https://www.legislation.gov.uk/ukpga/2018/12/contents}.
(Accessed: 20 April 2026).

\bibitem{icoResearchSafeguards}
Information Commissioner's Office (ICO).
\textit{The research provisions: appropriate safeguards under UK GDPR and DPA 2018}.
\url{https://ico.org.uk/for-organisations/uk-gdpr-guidance-and-resources/the-research-provisions/what-are-the-appropriate-safeguards}.
(Accessed: 20 April 2026).

\bibitem{icoDataMinimisation}
Information Commissioner's Office (ICO).
\textit{Principle (c): Data minimisation}.
\url{https://ico.org.uk/for-organisations/uk-gdpr-guidance-and-resources/data-protection-principles/a-guide-to-the-data-protection-principles/data-minimisation/}.
(Accessed: 20 April 2026).

\bibitem{bcsCodeConduct2022}
BCS, The Chartered Institute for IT (2022).
\textit{Code of Conduct for BCS Members (Version 8)}.
\url{https://www.bcs.org/media/2211/bcs-code-of-conduct.pdf}.
(Accessed: 20 April 2026).

\bibitem{bcsCodeGoodPractice1972}
British Computer Society (1972).
\textit{The British Computer Society Code of Good Practice}.
National Computing Centre.

\bibitem{zhang2022membership}
Zhang, Z., Yan, C., and Malin, B. A. (2022).
\textit{Membership Inference Attacks Against Synthetic Health Data}.
Journal of Biomedical Informatics, 125, 103977.

\bibitem{yan2022benchmarking}
Yan, C., Yan, Y., Wan, Z., Zhang, Z., Omberg, L., Guinney, J., Mooney, S. D., and Malin, B. A. (2022).
\textit{A multifaceted benchmarking of synthetic electronic health record generation models}.
Nature Communications, 13, 7609.

\bibitem{kaabachi2025scoping}
Kaabachi, B., Despraz, J., Meurers, T., Otte, K., Halilovic, M., Kulynych, B., Prasser, F., and Raisaro, J. L. (2025).
\textit{A scoping review of privacy and utility metrics in medical synthetic data}.
npj Digital Medicine, 8, 60.

\bibitem{ramachandranpillai2024btgan}
Ramachandranpillai, R., Sikder, M. F., Bergstrom, D., and Heintz, F. (2024).
\textit{Bt-GAN: Generating Fair Synthetic Healthdata via Bias-transforming Generative Adversarial Networks}.
Journal of Artificial Intelligence Research, 79.

\bibitem{obermeyer2019racialbias}
Obermeyer, Z., Powers, B., Vogeli, C., and Mullainathan, S. (2019).
\textit{Dissecting racial bias in an algorithm used to manage the health of populations}.
Science, 366(6464), 447--453.

\bibitem{pedregosa2011scikit}
Pedregosa, F., Varoquaux, G., Gramfort, A., Michel, V., Thirion, B., Grisel, O.,
Blondel, M., Prettenhofer, P., Weiss, R., Dubourg, V., Vanderplas, J., Passos, A.,
Cournapeau, D., Brucher, M., Perrot, M., and Duchesnay, E. (2011).
\textit{Scikit-learn: Machine Learning in Python}.
Journal of Machine Learning Research, 12, 2825--2830.

\bibitem{virtanen2020scipy}
Virtanen, P., Gommers, R., Oliphant, T. E., Haberland, M., Reddy, T., Cournapeau, D., et al. (2020).
\textit{SciPy 1.0: Fundamental Algorithms for Scientific Computing in Python}.
Nature Methods, 17(3), 261--272.

\bibitem{fawcett2006roc}
Fawcett, T. (2006).
\textit{An Introduction to ROC Analysis}.
Pattern Recognition Letters, 27(8), 861--874.

\bibitem{davisGoadrich2006}
Davis, J., and Goadrich, M. (2006).
\textit{The Relationship between Precision-Recall and ROC Curves}.
Proceedings of the 23rd International Conference on Machine Learning, 233--240.

\bibitem{pytestProject}
Krekel, H., Oliveira, B., Pfannschmidt, R., Bruynooghe, F., Laugher, B., Bruhin, F., et al.
\textit{pytest testing framework}.
\url{https://github.com/pytest-dev/pytest}.
(Accessed: 20 April 2026).

\bibitem{kotelnikov2023tabddpm}
Kotelnikov, A., Baranchuk, D., Rubachev, I., and Babenko, A. (2023).
\textit{TabDDPM: Modelling Tabular Data with Diffusion Models}.
Proceedings of the 40th International Conference on Machine Learning (ICML), PMLR 202, 17564--17579.

\bibitem{zhang2024tabsyn}
Zhang, H., Huang, J., Li, Z., Zhang, J., Gu, S., and Wu, Y. (2024).
\textit{Mixed-Type Tabular Data Synthesis with Score-based Diffusion in Latent Space}.
International Conference on Learning Representations (ICLR), Oral.

\bibitem{solatorio2023realtabformer}
Solatorio, A. V., and Dupriez, O. (2023).
\textit{REaLTabFormer: Generating Realistic Relational and Tabular Data using Transformers}.
arXiv:2302.02041.

\bibitem{borisov2023great}
Borisov, V., Sessler, K., Leemann, T., Pawelczyk, M., and Kasneci, G. (2023).
\textit{Language Models are Realistic Tabular Data Generators}.
International Conference on Learning Representations (ICLR).

\bibitem{kim2023stasy}
Kim, J., Lee, C., and Park, N. (2023).
\textit{STaSy: Score-based Tabular Data Synthesis}.
International Conference on Learning Representations (ICLR).

\bibitem{wang2022promptehr}
Wang, Z., and Sun, J. (2022).
\textit{PromptEHR: Conditional Electronic Healthcare Records Generation with Prompt Learning}.
Proceedings of the 2022 Conference on Empirical Methods in Natural Language Processing (EMNLP).

\bibitem{zhao2021ctabgan}
Zhao, Z., Kunar, A., Birke, R., and Chen, L. Y. (2021).
\textit{CTAB-GAN: Effective Table Data Synthesizing}.
Proceedings of the 13th Asian Conference on Machine Learning (ACML), PMLR 157, 97--112.
arXiv:2102.08369.

\bibitem{choi2017medgan}
Choi, E., Biswal, S., Malin, B., Duke, J., Stewart, W. F., and Sun, J. (2017).
\textit{Generating Multi-label Discrete Patient Records using Generative Adversarial Networks}.
Proceedings of the 2nd Machine Learning for Healthcare Conference (MLHC), PMLR 68, 286--305.
arXiv:1703.06490.

\bibitem{shi2025tabdiff}
Shi, J., Xu, M., Hua, H., Zhang, H., Ermon, S., and Leskovec, J. (2025).
\textit{TabDiff: a Multi-Modal Diffusion Model for Tabular Data Generation}.
International Conference on Learning Representations (ICLR).
arXiv:2410.20626.

\bibitem{song2021scoresde}
Song, Y., Sohl-Dickstein, J., Kingma, D. P., Kumar, A., Ermon, S., and Poole, B. (2021).
\textit{Score-Based Generative Modeling through Stochastic Differential Equations}.
International Conference on Learning Representations (ICLR), Oral.
arXiv:2011.13456.

\bibitem{tong2024cfm}
Tong, A., Malkin, N., Huguet, G., Zhang, Y., Rector-Brooks, J., Fatras, K., Wolf, G., and Bengio, Y. (2024).
\textit{Improving and generalizing flow-based generative models with minibatch optimal transport}.
Transactions on Machine Learning Research (TMLR).
arXiv:2302.00482.

\end{thebibliography}

\addcontentsline{toc}{section}{Bibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Appendices
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% List appendices in TOC only; full content is in lexisflow_appendices.pdf.
\addcontentsline{toc}{chapter}{Appendix A: Supplementary Sweep Results}
\addcontentsline{toc}{chapter}{Appendix B: User Guide}
\addcontentsline{toc}{chapter}{Appendix C: Source Code}
\end{document}
