\chapter{User Guide}

This appendix is the full operational reference for the pipeline (install, configuration, single-command sweeps, resumability, and result paths). The main report's Chapter~6 gives only the minimum needed to interpret methods and results; details that would duplicate the argumentative spine are concentrated here to avoid bloat in the main volume.
All commands assume a Unix-like terminal (macOS or Linux) and are run from the project root directory.

%==============================================================================
\section{Prerequisites}
%==============================================================================

\begin{itemize}\tightlist
    \item \textbf{Python 3.10 or later.}
    \item \textbf{uv} (recommended) or \textbf{pip} for dependency management. Install uv via:
    \begin{verbatim}
    curl -LsSf https://astral.sh/uv/install.sh | sh
    \end{verbatim}
    \item \textbf{MIMIC-III access.} A credentialed PhysioNet account with a signed data use agreement is required to obtain the source data. See \url{https://mimic.physionet.org/gettingstarted/access/}.
    \item \textbf{No credentialed access available?} If you cannot access the credentialed MIMIC-III dataset, use the public Challenge 2012 workflow documented in section~\ref{sec:challenge2012-sweep-quick-guide} instead.
    \item \textbf{Preprocessed MIMIC tables.} The pipeline expects hourly ICU tables produced under the preprocessing contract in section~\ref{sec:mimic-preprocess-contract}, placed under the project data directory.
\end{itemize}

%==============================================================================
\section{Installation}
%==============================================================================

Clone the repository and install dependencies:

\begin{verbatim}
git clone https://github.com/lianmatsuo/lexisflow
cd lexisflow
uv sync            # installs all dependencies from pyproject.toml
\end{verbatim}

Alternatively, with pip:

\begin{verbatim}
pip install -e .
\end{verbatim}

To verify the installation, run the test suite:

\begin{verbatim}
uv run pytest
\end{verbatim}

All unit tests should pass.

%==============================================================================
\section{Downloading the Source Data}
%==============================================================================

\subsection{MIMIC-III (Credentialed)}

\begin{enumerate}\tightlist
    \item Create or sign in to a PhysioNet account.
    \item Complete the required credentialing steps and data use agreement for MIMIC-III.
    \item Request access and, once approved, download the required MIMIC-III tables from PhysioNet:
    \url{https://physionet.org/content/mimiciii/1.4/}.
    \item Run the project preprocessing pipeline to produce the expected derived CSV inputs used by this repository (hour-0 and autoregressive tables).
\end{enumerate}

\subsection{PhysioNet Challenge 2012 (Public)}

If you do not have credentialed MIMIC-III access, use the public Challenge 2012 dataset:
\url{https://physionet.org/content/challenge-2012/1.0.0/}.

\begin{enumerate}\tightlist
    \item Download the Challenge 2012 training set files from the PhysioNet page above.
    \item Place raw files under the repository's Challenge 2012 data area (project \texttt{data/challenge2012/} tree).
    \item Run:
    \begin{verbatim}
uv run python scripts/run_sweep.py --dataset challenge2012 --reset --n-jobs 8
    \end{verbatim}
    This command prepares processed CSVs and executes the full Challenge 2012 sweep.
\end{enumerate}

%==============================================================================
\section{Configuration}
%==============================================================================

Runtime configuration is now centralised in Python dataset presets plus CLI overrides.
The authoritative defaults are defined in \texttt{src/lexisflow/config/datasets.py}.

\begin{itemize}\tightlist
    \item \textbf{Dataset paths (\texttt{DatasetConfig}):} script directory, processed CSV paths, preprocessor artifact paths, transformed cache location, results CSV, and sweep artifact directory.
    \item \textbf{Sweep defaults (\texttt{SweepDefaults}):} \texttt{nt} grid, \texttt{n\_noise} grid, training row cap, synthetic sample count, and privacy-evaluation row cap.
    \item \textbf{Split policy (\texttt{SplitConfig}):} patient-level test fraction, holdout fraction, and split shuffle seed.
    \item \textbf{Profiles (\texttt{sweep\_profiles}):} each dataset exposes \texttt{full} and \texttt{smoke} presets. Scripts resolve defaults from the selected profile, then apply explicit CLI overrides.
\end{itemize}

Practical rule: treat \texttt{datasets.py} as the single source of truth for default
paths, sweep budgets, and split ratios.

%==============================================================================
\section{Pipeline Walkthrough}
%==============================================================================

The pipeline is now executed through a single command that orchestrates data
preparation, preprocessor fitting, and the sweep run.

\subsection{Single Command Pipeline}

\begin{verbatim}
uv run python scripts/run_sweep.py --dataset mimic
uv run python scripts/run_sweep.py --dataset challenge2012
uv run python scripts/run_sweep.py --dataset mimic --reset
uv run python scripts/run_sweep.py --dataset mimic --profile smoke
uv run python scripts/run_sweep.py --dataset mimic --n-jobs 8
\end{verbatim}

\textbf{Outputs (MIMIC):}
\begin{itemize}\tightlist
    \item \texttt{data/processed/hour0\_data.csv}
    \item \texttt{data/processed/autoregressive\_data.csv}
    \item \texttt{artifacts/hour0\_preprocessor.pkl}
    \item \texttt{artifacts/preprocessor\_full.pkl}
    \item \texttt{results/sweep\_results.csv}
\end{itemize}

\textbf{Outputs (Challenge 2012):}
\begin{itemize}\tightlist
    \item \texttt{data/challenge2012/processed/hour0\_data.csv}
    \item \texttt{data/challenge2012/processed/autoregressive\_data.csv}
    \item \texttt{artifacts/challenge2012/hour0\_preprocessor.pkl}
    \item \texttt{artifacts/challenge2012/autoregressive\_preprocessor.pkl}
    \item \texttt{results/challenge2012\_sweep\_results.csv}
\end{itemize}

\subsection{Step 4: Evaluate Quality}

Quality metrics are computed automatically for every sweep cell (KS statistics, correlation Frobenius norm, clinical range violations, privacy diagnostics). Use \texttt{scripts/run\_sweep.py --dataset mimic} (or \texttt{--dataset challenge2012}) for end-to-end execution.

%==============================================================================
\section{Hyperparameter Sweep}
%==============================================================================

For systematic evaluation across configurations, use the unified sweep entrypoint:

\begin{verbatim}
uv run python scripts/run_sweep.py --dataset mimic
uv run python scripts/run_sweep.py --dataset challenge2012
uv run python scripts/run_sweep.py --dataset mimic --reset
\end{verbatim}

Key options:
\begin{itemize}\tightlist
    \item \texttt{--dataset}: choose \texttt{mimic} or \texttt{challenge2012}.
    \item \texttt{--profile}: choose \texttt{full} or \texttt{smoke}; forwarded to the dataset-specific sweep driver.
    \item \texttt{--n-jobs}: override CPU worker count used during the sweep stage.
    \item \texttt{--reset}: clear generated data, preprocessors, cache and sweep outputs before rerun.
\end{itemize}

The sweep trains both hour-0 and autoregressive models for each
$(n_t, n_{\text{noise}})$ configuration, generates synthetic samples, runs
multi-task TSTR and quality/privacy metrics, and appends to the dataset-specific
results CSV (\texttt{results/sweep\_results.csv} for MIMIC,
\texttt{results/challenge2012\_sweep\_results.csv} for Challenge 2012). The
sweep is resumable: existing rows are skipped on rerun.

\subsection{Challenge 2012 Sweep Quick Guide}
\label{sec:challenge2012-sweep-quick-guide}

Use this sequence when you specifically want the public-benchmark run:
\begin{enumerate}\tightlist
    \item \textbf{Fresh run (recommended first pass):}
    \begin{verbatim}
uv run python scripts/run_sweep.py --dataset challenge2012 --reset --n-jobs 8
    \end{verbatim}
    This rebuilds processed CSVs, preprocessors, transformed cache, model artifacts, and
    \texttt{results/challenge2012\_sweep\_results.csv}.
    \item \textbf{Resume without deleting prior work:}
    \begin{verbatim}
uv run python scripts/run_sweep.py --dataset challenge2012 --n-jobs 8
    \end{verbatim}
    Existing \texttt{(nt, n\_noise)} rows are skipped automatically.
    \item \textbf{Low-cost smoke validation:}
    \begin{verbatim}
uv run python scripts/run_sweep.py --dataset challenge2012 --profile smoke
    \end{verbatim}
\end{enumerate}

\paragraph{Important CLI distinction.}
\texttt{--reset} is a flag of the \emph{top-level} driver
(\texttt{scripts/run\_sweep.py}), not the dataset-specific
\texttt{scripts/challenge2012/run\_sweep.py}. Use the top-level command when
you want artifact cleanup before rerun.

\paragraph{How to read Challenge 2012 outputs.}
The main file is \texttt{results/challenge2012\_sweep\_results.csv}. In this
CSV:
\begin{itemize}\tightlist
    \item \texttt{error} contains failure messages for interrupted/failed cells; successful rows have an empty error field.
    \item \texttt{los\_synth\_roc\_auc} and \texttt{los\_real\_roc\_auc} are valid when the run uses the updated LOS ROC-AUC fallback path for two-class splits.
    \item \texttt{stay\_length\_ks} should be interpreted carefully on Challenge 2012 because both real and synthetic tables are represented as fixed 48-hour windows in this pipeline.
    \item \texttt{range\_violation\_pct} can be influenced by source-data outliers (not only generator error), especially in blood-pressure fields.
\end{itemize}

Dataset sweep drivers (\texttt{scripts/mimic/run\_sweep.py} and
\texttt{scripts/challenge2012/run\_sweep.py}) also support direct execution with
fine-grained overrides:
\begin{itemize}\tightlist
    \item \texttt{--profile}, \texttt{--train-rows}, \texttt{--synth-samples}
    \item \texttt{--nt-values}, \texttt{--noise-values}, \texttt{--n-jobs}
    \item \texttt{--model-type} (\texttt{hs3f}, \texttt{forest-flow}, \texttt{ctgan})
    \item \texttt{--refresh-transformed-cache}, \texttt{--skip-privacy-metrics}, \texttt{--privacy-max-rows}
\end{itemize}

To visualise sweep results:

\begin{verbatim}
uv run python scripts/common/analyze_sweep.py
uv run python scripts/common/analyze_sweep.py --dataset challenge2012
uv run python scripts/common/analyze_sweep.py --results-path results/challenge2012_sweep_results.csv
\end{verbatim}

\texttt{analyze\_sweep.py} options:
\begin{itemize}\tightlist
    \item \texttt{--dataset}: selects default results path from dataset config.
    \item \texttt{--results-path}: explicit input CSV override.
    \item \texttt{--output-dir}: explicit plot output directory override.
\end{itemize}

%==============================================================================
\section{Running Tests}
%==============================================================================

The project includes 54 unit tests covering quality metrics, privacy metrics, TSTR framework, autoregressive transformation, feature utilities, model training, and sampling:

\begin{verbatim}
uv run pytest                    # run all tests
uv run pytest -v                 # verbose output
uv run pytest src/lexisflow/models/   # run model tests only
\end{verbatim}
